UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2018-12-07 ms/main.tex
rm comments, details in umap vs tsne pgh

Uniform Manifold Approximation and Projection (UMAP) is a new dimension reduction technique grounded in Riemannian geometry, algebraic topology, and category theory, and designed to model and preserve the high-dimensional topology of data points in the low-dimensional space\cite{2018arXivUMAP}. The assumption behind UMAP is that data are uniformly distributed on local manifolds in high-dimensional space, which can be approximated as fuzzy sets that are patched together to form a topological representation. One can then construct a low-dimensional topological representation that minimizes the differences between the two representations. With genotype data, UMAP creates a neighbourhood around each individual's genetic coordinates and identifies a pre-selected number of neighbours to build high-dimensional manifolds. The end result is a low-dimension representation that groups genetically similar individuals together on a local scale while preserving long-range topological connections to more distantly related individuals.\adpcomment{added this sentence.} The method has been successfully applied to single-cell RNA sequencing datasets\cite{umap2018singlecell}.

@@

To compare UMAP to t-SNE directly and assess the role of convergence, we ran UMAP and Python's sklearn implementation of t-SNE, and a hybrid approach where the early exaggeration phase of t-SNE (typically the first 250 iterations) has been replaced by a UMAP run with \sgcomment{x epochs}. The three approaches produce similar results in smaller datasets such as the 1KGP and HRS, with the main differences being in run-time and some qualitative factors such as clusters being more interconnected when using UMAP (figures~\ref{fig:supp_tsne_umap_compare_1kgp},~\ref{fig:supp_tsne_umap_compare_hrs}). When applied to the UKbiobank, we find (perhaps subjectively) that UMAP provides a more compelling visualization of the data compared to both version of t-SNE. We also find that the hybrid approach produces a better value for the t-SNE metric than the default t-SNE approach (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}), confirming that the default t-SNE optimization fails to identify the global t-SNE minimum. Qualitatively, the standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but broke the global continuity between clusters (figure~\ref{fig:supp_tsne_umap_compare_ukbb}). Given the many desirable properties of UMAP, such as computational performance and intuitive placement of individuals within a global genetic continuum, we conclude that UMAP is a superior alternative to t-SNE for large and diverse genomic datasets.

>>>>

Uniform Manifold Approximation and Projection (UMAP) is a new dimension reduction technique grounded in Riemannian geometry, algebraic topology, and category theory, and designed to model and preserve the high-dimensional topology of data points in the low-dimensional space\cite{2018arXivUMAP}. The assumption behind UMAP is that data are uniformly distributed on local manifolds in high-dimensional space, which can be approximated as fuzzy sets that are patched together to form a topological representation. One can then construct a low-dimensional topological representation that minimizes the differences between the two representations. With genotype data, UMAP creates a neighbourhood around each individual's genetic coordinates and identifies a pre-selected number of neighbours to build high-dimensional manifolds. The end result is a low-dimension representation that groups genetically similar individuals together on a local scale while preserving long-range topological connections to more distantly related individuals. The method has been successfully applied to single-cell RNA sequencing datasets\cite{umap2018singlecell}.

@@

To compare UMAP to t-SNE directly and assess the role of convergence, we ran UMAP and Python's sklearn implementation of t-SNE, and a hybrid approach where the early exaggeration phase of t-SNE (typically the first 250 iterations) has been replaced by a full UMAP run (600 epochs for the 1KGP data and 230 epochs for the HRS and UKBB data). The three approaches produce similar results in smaller datasets such as the 1KGP and HRS, with the main differences being in run-time and some qualitative factors such as clusters being more interconnected when using UMAP (figures~\ref{fig:supp_tsne_umap_compare_1kgp},~\ref{fig:supp_tsne_umap_compare_hrs}). When applied to the UKbiobank, we find (perhaps subjectively) that UMAP provides a more compelling visualization of the data compared to both version of t-SNE. We also find that the hybrid approach produces a better value for the t-SNE metric than the default t-SNE approach (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}), confirming that the default t-SNE optimization fails to identify the global t-SNE minimum. Qualitatively, the standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but broke the global continuity between clusters (figure~\ref{fig:supp_tsne_umap_compare_ukbb}). An animation of t-SNE iterations starting from the UMAP embedding can be seen in the supplementary files. Given the many desirable properties of UMAP, such as computational performance and intuitive placement of individuals within a global genetic continuum, we conclude that UMAP is a superior alternative to t-SNE for large and diverse genomic datasets.
